Hydraulic bore interaction with a column - A comparison between the solution of the shallow equation and experimental results

### 2015-03-20 doc/paper/paper.tex
correct some grammar error according to comment by Randy

\begin{abstract}
In this paper we compare the solution of the shallow water equations with experimental results. The experiment simulates the interaction between an incident bore and a free-standing coastal structure. The shallow water equations are solved using the GeoClaw solver of the CLAWPACK software.
\par
We simulate and compare three different cases. In the first case we performe and evaluate a simple dam break problem.
We compare the history of the wave height and velocity at the center location, where the column will be mounted. 
In the second case we add a square column and compare wave height at various locations in front of and behind the column. 
For the third case we replace the square column by a cylindrical column and compare the same variables with the experimental results. 
We analyze and discuss the discrepancy between the GeoClaw simulation and the experimental results.
\end{abstract}

@@

            \caption{Diagram of experiment apparatus}

@@

In all cases below, all boundary conditions are set to reflecting walls. 
\section{Dam Break}\label{Sec:Zero}
The initial condition for this and the following cases is given by
\eqm
h_0(x,y)=& 
\begin{cases}
0.25 & x<5.9 \label{eq:init1}\\
0.02 & \text{otherwise}
\end{cases}\\
u_0(x,y) =& 0\label{eq:init2}
\enm
In this comparison we will include results from OpenFOAM, an open source CFD software modeling air and water using the Navier-Stokes equations with turbulence model.
\par
If we analyze the Rankine-Hugoniot condition of the initial conditions (\ref{eq:init1}) and (\ref{eq:init2}) for the shallow water equations:
\eqm
s(h_*-h) &=& h_*u_* - hu\\
s(h_*u_*-hu) &=& hu_*^2-hu^2 +\frac{1}{2}g(h_*^2-h^2)
\enm
we see that we can not find a $s$ that satisfies both equations, since $u$ and $u*$ are zero. This means the Rankine-Hugoniot condition is not satisfied and we expect a right and left going wave originating form the dam break.
Figure \ref{fig:waveheight_nocolumn} shows the time history of the wave height, measured at the center location where the column will be mounted in the following cases. 
The result from GeoClaw agrees quite well in general. 
If we zoom in Figure \ref{fig:waveheight_nocolumn} for smaller time range from $t=3s$ to $t=4s$, we get Figure \ref{fig:waveheight_nocolumn_zoomin}.
In Figure \ref{fig:waveheight_nocolumn_zoomin}, GeoClaw accurately predicted the arrival of the shock at t=3.8s. The arrival time of peak predicted by GeoClaw is almost the same as that from experiment (only about 0.02 second earlier).  
GeoClaw overestimates the peak water level a little, while OpenFOAM gives a better value.
At a later time, the wave gets reflected at the right wall of the tank, which can be seen by the second jump in wave height. 
GeoClaw also predicts the arrival time for this peak quite well, but overestimates the peak value by a small amount.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_nocolumn}
        \caption{time from 0 to 18}
        \label{fig:waveheight_nocolumn}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_nocolumn_zoomin}
        \caption{time from 3 to 4}
        \label{fig:waveheight_nocolumn_zoomin}
    \end{subfigure}
    \caption{time history of wave height for case without a column. }
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{./plots/velocityU_nocolumn}
    \caption{time history of streamwise velocity U for case without a column}
    \label{fig:velocityU_nocolumn}
\end{figure}
Figure \ref{fig:velocityU_nocolumn} shows the time history of the streamwise velocity $u$ from GeoClaw and experiment. 
Note that this velocity $u$ in both GeoClaw and experiment are depth-averaged. 
GeoClaw accurately predicted the time of the velocity peak as well as the value of the peak. 
After the initial jump in the velocity, GeoClaw overestimated the value of velocity.
The reason why GeoClaw gives a higher velocity may be attributed to its depth-averaged property: water at all depth are propagating with the same speed. In the experiment, the velocity is getting smaller the closer we get to the bottom.

@@

This difference indicates that the column gives more resistance to shock waves in experiment than it does in shallow water equation model. 

@@

The shallow water equations are essentially a system of hyperbolic PDEs. The solution depends on initial condition along characteristics in the spatio-temporal domain. 

@@

In real physics, or in Navier-Stokes equations, when front of the shock wave gets close to the column, water downstream it and upstream column are affected (have nonzero velocity) and resist the comming of wave to some extent. 
In real world this slows down the propagation speed of the shock wave when it comes close to the column and thus shifts the peak toward later time in the plots.

@@

        \caption{at x = 11.0 (0.04m ahead of front face of the square column)}

@@

        \caption{at x = 11.2 (0.02m ahead of front face of the square column)}

@@

    \caption{time history of wave height for case with a square column.}

@@

        \caption{comparison between case with a square column and case without one}

@@

        \caption{case with a square column with a larger time range at x = 11.04}

@@

    \caption{time history of wave height for square column case}

@@

Figure \ref{fig:waveheight_square_back} shows history of wave height measured at back face of the square column (at $X = 11.16$). Boundary layer detached near this region thus fluid in this region is difficult to predict. GeoClaw overestimates wave height in this region.  

@@

    \caption{history of wave height measure at back face of the square column}

@@

\section{Dam Break with cylindrical column}\label{Sec:Cylinder}
Figure \ref{fig:waveheight_cylinder_x=11.0} shows the time history of the wave height measured at $0.04m$ ahead of the front point of cylindrical column($0.1m$ ahead of the center of the column).
%Since the way GeoClaw treat a cell partly belong to cylinder and partly belong to tank bottom is to give that cell an average value of topography from cylinder and tank bottom based on area ratio, we get a quasi-circle actually instead of circle. Thus the result in this case does not agree well with the experiment result. 
%Future test can be conducted by setting a higher level for mesh refinement to get a better circle.
Similar feature as in the square column case Figure \ref{fig:waveheight_square_x=11.0} can be found.
We can see a decrease in peak value for this case compared with the same plots for square column case. However, this decrease between two cases from GeoClaw is not as large as that between two cases from experiment. 
\par
Note that diameter of the cylinder is equal to half the length of the square column. 
Thus in experiment, the square column can block the flow in in a more efficient way than cylinder does, since the cylinder has a smoother shape for the fluid to pass by. 
This explains why we see a lower peak of wave height, as well as less steep curve for history of wave height in the case with the cylinder in the experiment. 
In GeoClaw, a geometry is represented with quadrilateral cells. 
In the cylinder case, the cylinder is represented with square cells since our background mesh is simple structured quadrilateral mesh with square cells.
Based on this, we can say that in GeoClaw, when the shock wave hits the cylinder, it is hitting onto many small faces vertical to the flow direction at different locations. 
In other words, there is no cell which has a normal vector of front face pointing toward oblique direction other than the direction the fluid comes.  
Thus, it is similar to hitting onto one big face vertical to the flow direction, which is the case with a square column.  
This is why we get similar results in GeoClaw for the case with square column and the case with a cylinder.
Figure \ref{fig:waveheight_square_x=11.0} also gives results after we refine the mesh. As the mesh becomes finer, the peak value decrease as we expect.
But with a even finer mesh, we cannot decrease the peak value, which indicates that a mapped grid could be used to improve the solution.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x11_finer_new}
        \caption{at x = 11}
        \label{fig:waveheight_cylinder_x=11.0}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x1104_largerTimeScale}
        \caption{at x = 11.04}
        \label{fig:waveheight_cylinder_x=11.04}
    \end{subfigure}
    \caption{time history of wave height for cylindrical column case}
\end{figure}
\par
Figure \ref{fig:waveheight_cylinder_back} shows history of wave height measured at back face of the cylinder column (at $X = 11.16$). 
GeoClaw gives a satisfactory prediction in this region.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{./plots/waveheight_cylinder_back}
    \caption{history of wave height measure at back face of the square column}
    \label{fig:waveheight_cylinder_back}
\end{figure}

>>>>

\begin{abstract}
In this paper we compare the solution of the shallow water equations with experimental results. The experiment simulates the interaction between an incident bore and a free-standing coastal structure. The shallow water equations are solved using the GeoClaw solver of the CLAWPACK software.
\par
We simulate and compare three different cases. In the first case we perform and evaluate a simple dam break problem.
We compare the history of the wave height and velocity at the center location, where the column will be mounted. 
In the second case we add a square column and compare wave height at various locations in front of and behind the column. 
For the third case we replace the square column by a cylindrical column and compare the same variables with the experimental results. 
We analyze and discuss the discrepancy between the GeoClaw simulation and the experimental results.
\end{abstract}

@@

            %\caption{Diagram of experiment apparatus}
            \caption{Plan view and side view of experiment apparatus}

@@

In all cases below, all boundary conditions are set to reflecting walls. 
\section{Dam Break}\label{Sec:Zero}
The initial condition for this and the following cases is given by
\eqm
h_0(x,y)=& 
\begin{cases}
0.25 & x<5.9 \label{eq:init1}\\
0.02 & \text{otherwise}
\end{cases}\\
u_0(x,y) =& 0\label{eq:init2}
\enm
In this comparison we will include results from OpenFOAM, an open source CFD software modeling air and water using the Navier-Stokes equations with turbulence model.
\par
If we analyze the Rankine-Hugoniot condition of the initial conditions (\ref{eq:init1}) and (\ref{eq:init2}) for the shallow water equations:
\eqm
s(h_*-h) &=& h_*u_* - hu\\
s(h_*u_*-hu) &=& hu_*^2-hu^2 +\frac{1}{2}g(h_*^2-h^2)
\enm
we see that we can not find a $s$ that satisfies both equations, since $u$ and $u*$ are zero. This means the Rankine-Hugoniot condition is not satisfied and we expect a right and left going wave originating form the dam break.
Figure \ref{fig:waveheight_nocolumn} shows the time history of the wave height, measured at the center location where the column will be mounted in the following cases. 
The result from GeoClaw agrees quite well in general. 
If we zoom in Figure \ref{fig:waveheight_nocolumn} for smaller time range from $t=3s$ to $t=4s$, we get Figure \ref{fig:waveheight_nocolumn_zoomin}.
In Figure \ref{fig:waveheight_nocolumn_zoomin}, GeoClaw accurately predicted the arrival of the shock at t=3.8s. The arrival time of peak predicted by GeoClaw is almost the same as that from experiment (only about 0.02 second earlier).  
GeoClaw overestimates the peak water level a little, while OpenFOAM gives a better value.
At a later time, the wave gets reflected at the right wall of the tank, which can be seen by the second jump in wave height. 
GeoClaw also predicts the arrival time for this peak quite well, but overestimates the peak value by a small amount.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_nocolumn}
        \caption{Time from 0 to 18}
        \label{fig:waveheight_nocolumn}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_nocolumn_zoomin}
        \caption{Time from 3 to 4}
        \label{fig:waveheight_nocolumn_zoomin}
    \end{subfigure}
    \caption{Time history of wave height for case without a column. }
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{./plots/velocityU_nocolumn}
    \caption{Time history of streamwise velocity U for case without a column}
    \label{fig:velocityU_nocolumn}
\end{figure}
Figure \ref{fig:velocityU_nocolumn} shows the time history of the streamwise velocity $u$ from GeoClaw and experiment. 
Note that this velocity $u$ in both GeoClaw and experiment are depth-averaged. 
GeoClaw accurately predicted the time of the velocity peak as well as the value of the peak. 
After the initial jump in the velocity, GeoClaw overestimated the value of velocity.
The reason why GeoClaw gives a higher velocity may be attributed to its depth-averaged property: water at all depth are propagating with the same speed. In the experiment, the velocity is getting smaller the closer we get to the bottom.

@@

This difference indicates that the column gives more resistance to shock waves in the experiment than it does in the shallow water equation model. 

@@

The shallow water equations are essentially a system of hyperbolic PDEs. The solution depends on initial conditions along characteristics in the spatio-temporal domain. 

@@

In real physics, or in Navier-Stokes equations, when the front of the shock wave gets close to the column, water downstream it and upstream column are affected (have nonzero velocity) and resist the coming of wave to some extent. 
In the real world this slows down the propagation speed of the shock wave when it comes close to the column and thus shifts the peak toward later time in the plots.

@@

        \caption{At x = 11.0 (0.04m ahead of front face of the square column)}

@@

        \caption{At x = 11.2 (0.02m ahead of front face of the square column)}

@@

    \caption{Time history of wave height for the case with a square column.}

@@

        \caption{Comparison between case with a square column and case without one}

@@

        \caption{Case with a square column with a larger time range at x = 11.04}

@@

    \caption{Time history of wave height for square column case}

@@

Figure \ref{fig:waveheight_square_back} shows history of wave height measured at back face of the square column (at $X = 11.16$). The boundary layer is detached near this region and thus fluid in this region is difficult to predict. GeoClaw overestimates wave height in this region.  

@@

    \caption{History of wave height measured at the back face of the square column}

@@

\section{Dam Break with cylindrical column}\label{Sec:Cylinder}
Figure \ref{fig:waveheight_cylinder_x=11.0} shows the time history of the wave height measured at $0.04m$ ahead of the front point of a cylindrical column($0.1m$ ahead of the center of the column).
%Since the way GeoClaw treat a cell partly belong to cylinder and partly belong to tank bottom is to give that cell an average value of topography from cylinder and tank bottom based on area ratio, we get a quasi-circle actually instead of circle. Thus the result in this case does not agree well with the experiment result. 
%Future test can be conducted by setting a higher level for mesh refinement to get a better circle.
Similar features as in the square column case Figure \ref{fig:waveheight_square_x=11.0} can be found.
We can see a decrease in peak value for this case compared with the same plots for the square column case. However, this decrease between two cases from GeoClaw is not as large as that between two cases from experiment. 
\par
Note that diameter of the cylinder is equal to half the length of the square column. 
Thus in experiment, the square column can block the flow in in a more efficient way than cylinder does, since the cylinder has a smoother shape for the fluid to pass by. 
This explains why we see a lower peak of wave height, as well as less steep curve for history of wave height in the case with the cylinder in the experiment. 
In GeoClaw, a geometry is represented with quadrilateral cells. 
In the cylinder case, the cylinder is represented with square cells since our background mesh is simple structured quadrilateral mesh with square cells.
Based on this, we can say that in GeoClaw, when the shock wave hits the cylinder, it is hitting onto many small faces vertical to the flow direction at different locations. 
In other words, there is no cell which has a normal vector of front face pointing toward oblique direction other than the direction the fluid comes.  
Thus, it is similar to hitting onto one big face vertical to the flow direction, which is the case with a square column.  
This is why we get similar results in GeoClaw for the case with square column and the case with a cylinder.
Figure \ref{fig:waveheight_square_x=11.0} also gives results after we refine the mesh. As the mesh becomes finer, the peak value decrease as we expect.
But with a even finer mesh, we cannot decrease the peak value, which indicates that a mapped grid could be used to improve the solution.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x11_finer_new}
        \caption{At x = 11}
        \label{fig:waveheight_cylinder_x=11.0}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x1104_largerTimeScale}
        \caption{At x = 11.04}
        \label{fig:waveheight_cylinder_x=11.04}
    \end{subfigure}
    \caption{Time history of wave height for cylindrical column case}
\end{figure}
\par
Figure \ref{fig:waveheight_cylinder_back} shows history of wave height measured at back face of the cylinder column (at $X = 11.16$). 
GeoClaw gives a satisfactory prediction in this region.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{./plots/waveheight_cylinder_back}
    \caption{History of wave height measure at back face of the square column}
    \label{fig:waveheight_cylinder_back}
\end{figure}
